# Materials and Methods

## Data

### DHS Data

The 2024 Nigeria DHS survey (NDHS) was conducted between December 2023 and May 2024 by the National Population Commission (NPC) and was designed to provide nationally representative estimates for Nigeria as a whole, as well as for urban and rural areas, the six geopolitical zones, and each of the 36 states and the Federal Capital Territory (FCT). The 2024 NDHS employed a stratified two-stage cluster sampling design based on the updated cartographic sampling frame prepared for Nigeria’s Population and Housing Census. The sampling frame excluded institutional populations such as persons in hotels, barracks, and prisons. In the first stage, enumeration areas (EAs), which served as primary sampling units or clusters, were selected with probability proportional to size within each of the 74 sampling strata defined by state and place of residence (urban/rural). A total of 1,400 clusters were selected, comprising 701 urban and 699 rural clusters. In the second stage, a household listing was conducted in each selected cluster, after which 30 households were selected using systematic sampling, yielding the intended sample of approximately 42,000 households[@ndhs].

Data for this study were obtained from the Women's Questionnaire of the 2024 NDHS. Women aged 15–49 years who were usual members of selected households or who had spent the night preceding the survey in the selected household were eligible for interview. The questionnaire collected information on respondents' sociodemographic characteristics, reproductive history, maternal and child health, family planning, and other health-related indicators. In total, 39,553 women were eligible for interview (19,166 from urban areas and 20,387 from rural areas), and a response rate of 99% was achieved nationally as well as in both urban and rural areas.

Information on childhood vaccination was obtained for children born in the three years preceding the survey, and alive at the time of the survey. For each eligible child, vaccination information, including receipt of each vaccine antigen and the corresponding date of administration, was extracted from the child's vaccination card where available. In the absence of a vaccination card, vaccination status was determined based on maternal recall. According to the 2024 NDHS, vaccination cards were seen for 50.2% of children aged 12–23 months and 34.4% of children aged 24–35 months, indicating that maternal recall constituted an important source of vaccination information for a substantial proportion of children[@ndhs].

### Temperature data

Data on daily mean 2-metre air temperature for the period 1940–2024 and daily mean 2-metre dewpoint temperature for the period 2020–2024 were obtained from the European Centre for Medium-Range Weather Forecasts (ECMWF) ERA5 reanalysis via the Climate Data Store[@hersbach2023; @copernicus2023]. The data are provided on a regular 0.25° × 0.25° latitude–longitude grid (approximately 27.8 kilometres squared). The data were extracted for a spatial bounding box enclosing Nigeria (west: 2.676932°E, south: 4.0690959°N, east: 14.678014°E, north: 13.885645°N) and retrieval was carried out programmatically in R using the `ecmwfr` package[@hufkens2019], which provides an interface to the ECMWF data services and the Climate Data Store API. The downloaded files were stored on secure local storage in NetCDF format, with one file saved for each year to facilitate efficient management and analysis.

## Study population

The analysis was restricted to children born in the three years preceding the 2024 Nigeria Demographic and Health Survey. Children who had died before the survey were excluded since the DHS birth recode only collects vaccination information for living children and children who had missing information on date of birth were also excluded. For each vaccine antigen, the analysis included only children who had reached the recommended age of eligibility for that vaccine at the time of interview. For example, analyses of the first dose of the measles-containing vaccine (MCV1), which is scheduled at 9 months of age, were restricted to children aged at least 9 months. Therefore, the analytical sample differed slightly across vaccine antigens.

## Outcome variables

The primary outcomes were delays in receipt of five key routinely recommended childhood vaccine antigens: Bacillus Calmette–Guérin (BCG), the first, second and third doses of the pentavalent vaccine (Pentavalent 1, Pentavalent 2 and Pentavalent 3), Vitamin A, and the first dose of the measles-containing vaccine (MCV1). The pentavalent vaccine protects against the combination of diphtheria, tetanus, pertussis, hepatitis B and Haemophilus influenzae type b infections. According to Nigeria's Expanded Programme on Immunization (EPI) schedule[@ophori2014], BCG is recommended at birth, Pentavalent 1 at 6 weeks, Pentavalent 2 at 10 weeks, Pentavalent 3 at 14 weeks, vitamin A at 6 months, and MCV1 at 9 months. These antigens were selected because they represent key milestones in the infant immunization schedule and are widely used as indicators of the performance of routine immunization programmes during the first year of a child's life.

For each antigen, the recommended vaccination date was calculated by adding the scheduled age at vaccination to the child's date of birth. Timeliness was assessed at 7 days, 14 days, 21 days and 28 days. Although no universally accepted definition of vaccination delay exists [@masters2019], this study adpots the World Health Organization (WHO) definition of vaccine delay as receipt of a vaccine more than 28 days after its recommended due date [@worldhealthorganization2021], so that vaccines administered within 28 days of the scheduled date were classified as timely. This definition was adopted across all the antigens considered for analysis

## Exposure variables

The data on daily mean 2-metre air temperature and mean 2-metre dewpoint temperature from ERA5 were used to derive heatwave and heat index exposure variables. The ERA5 temperature data were spatially linked to the 2024 Nigeria DHS clusters using the geographic coordinates provided in the survey. Following DHS recommendations for incorporating spatial covariates while accounting for the random displacement of cluster coordinates [@perez-heydrich2013] (The DHS applies random displacement to cluster coordinates to protect confidentiality, with displacement distances of up to 2 km for urban clusters, 5 km for 99% of rural clusters, and 10 km for 1% of rural clusters). In this study, cluster locations were matched to ERA5 grid cells using a 10-kilometre buffer radius. Given that the ERA5 spatial resolution is approximately 0.25° (approximately 27 km), which is substantially larger than the maximum DHS displacement distance, coordinate displacement would be expected to have limited influence on the assigned temperature exposure. Therefore, the spatial linkage would generally correspond to assigning each DHS cluster the temperature series from the ERA5 grid cell containing its displaced centroid. Following this spatial linkage, all subsequent analyses were conducted using temperature time series corresponding to each DHS cluster location.

### Heatwave exposure variable

Heatwaves were defined following the probabilistic extreme event attribution framework [@philip2020; @clarke2025; @vanoldenborgh2022]. Specifically, a heatwave was defined as a 5-day rolling average temperature exceeding the estimated local 75th percentile derived from a fitted GEV model.

To capture sustained periods of unusually high temperatures and account for the cumulative effects of heat exposure, a five-day rolling mean of daily 2-metre air temperature (Tx5x index) was calculated for each DHS cluster location for the period January 1940 to May 2024. The resulting temperature time series was divided into a baseline period (January 1940–December 2019) and an analysis period (January 2020–May 2024). The baseline period was used to estimate cluster-specific extreme temperature thresholds, while the analysis period was used to identify heatwave events occurring during the period covered by the NDHS. For each DHS cluster, the annual maximum of the Tx5x index was extracted from each year of the baseline period, yielding 80 annual maxima. These annual maxima were modelled using a non-stationary Generalized Extreme Value (GEV) distribution in which the location parameter varied linearly with the Global Mean Surface Temperatures (GMST) to account for long-term warming trends[@philip2020; @vanoldenborgh2022]:

$$M_{t,x} \sim GEV(\mu, \sigma, \zeta)$$ $$\mu=\beta_0+\beta_1\cdot GMST$$ where $M_{t,x}$ denotes the annual maximum Tx5x value at cluster $x$ in year $t$, and $GMST$ denotes the Global Mean Surface Temperature in year $t$. The model was fitted using the `gamlss` and `gamlssx` packages in R[@r.a.rigby2005; @northrop2025].

The fitted GEV models were subsequently used to estimate $Z_4$, the 4-year return level for each DHS cluster, representing the temperature expected to be exceeded, on average, once every four years under the fitted non-stationary climate model. If $F(\cdot)$ denotes the fitted GEV cumulative distribution function[@rigby2019]:

$$F(x)=exp(-[1+\zeta(\frac{x-\mu}{\sigma})]^{-1/\zeta}_+)$$ Then, the $T$-year return level satisfies

$$P(M_{t,x}>Z_T)=\frac{1}{T}; \\F(Z_T)=1-\frac{1}{T}$$

For a 4-year return level, $F(Z_{4})=1-1/4=0.75$ and hence $Z_{4}=F^{-1}(0.75)$ which is given by:

$$Z_{4}=\mu+\frac{\sigma}{\zeta}[(-\log_e(0.75))^{-\zeta}-1], \zeta\neq0$$

Each Tx5x observation during the analysis period (January 2020–May 2024) was then compared with the corresponding DHS cluster-specific return level. Days for which the Tx5x index exceeded the return level $(Z_{4})$ were classified as heatwave days, whereas all remaining days were classified as non-heatwave days. This produced a daily binary heatwave indicator for every DHS cluster and each observation was classified as heatwave-exposed if at least one heatwave day occurred within a 15-day exposure window centered on the scheduled vaccination date (7 days before to 7 days after).

### Heat Index exposure variable

The heat index was calculated using 2-metre air temperature and 2-metre dewpoint temperature. The calculation followed the heat index algorithm developed by the U.S. National Weather Service (NWS), which is described by Anderson et al. [@anderson2013; @nationalweatherservice2011]. The weathermetrics package [@anderson2013] was used to implement this algorithm.

The heat index was categorized according to NWS guidelines as shown:

```{r}
#| echo: false
#| message: false
#| warning: false

pacman::p_load(kableExtra, knitr)

data.frame(
  Classification = c("Tolerable", "Caution", "Extreme Caution", "Danger", "Extreme Danger"),
  `Heat Index (°C)` = c("<26.7°C", "26.7°C–32.2°C", "32.2°C–39.4°C", "39.4°C–51.1°C", "≥51.7°C"),
  check.names = FALSE
) |>
  kbl(booktabs = TRUE, align = "lc", escape = FALSE) |>
  kable_styling(full_width = FALSE, latex_options = "hold_position") |>
  row_spec(0, bold = TRUE)
```

## Covariates

In this study, the covariates utilized were informed by a recent systematic review of factors associated with vaccine timeliness in low- and middle-income countries over 10 years (2007-2017)[@masters2019], and availability of the covariates in the DHS data. These comprised child, maternal, household and geographical characteristics that have previously been associated with vaccination timeliness. Continuous variables were categorized where appropriate to improve interpretability and to ensure consistency with standard DHS reporting. Maternal age at the child's birth was grouped into seven five-year age categories. Place of residence, place of delivery, child's sex, maternal education, household wealth index, perceived travel time to the nearest health facility, and geographical zone were analysed using the standard DHS classifications. Birth order was retained as a continuous variable. The covariates included in the analysis and their corresponding types are presented in Table 3.1.

```{r, t1}
#| tbl-pos: "H"
#| echo: false
#| message: false
#| warning: false

library(kableExtra)
library(knitr)

data.frame(
  `Variable` = c(
    'Gender', 'Delivery', 'Birth order', 
    'Education', 'Age-at-birth', 'Geozone',
    'DHF', 'Residence', 'Wealth'
    ), 
  `Description` = c(
    "The child's gender", "The place of delivery categorized as either institutional or home delivery", 
    "The child's birth order", "The highest level of education attained by the mother. Categories include: No education, primary, secondary and higher education.",
    "The age of the mother at birth", "Geopolitical zone. Nigeria is divided into six of these zones: North-West, North-East, North-Central (Middle Belt), South-West, South-East, and South-South", 
    "The distance to the nearest health facility in minutes", "The residence, categorized as rural or urban",
    "The household wealth index categorized into 5 quintiles (poorest, poorer, middle, richer, richest"
  ),
  `Type` = c(
    "Binary", "Binary", "Integer", "Categorical", "Categorical", "Categorical", "Categorical",
    "Binary", "Categorical"
  )
) |>
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Description of predictor variables included in the analysis.",
    align = c("l", "p{9cm}", "c"),
    linesep = ""
  ) |>
  kable_styling(latex_options = "hold_position") |>
  # kable_styling(
  #   latex_options = c("hold_position", "striped"),
  #   font_size = 10,
  #   full_width = FALSE,
  #   position = "center"
  # ) |>
  # column_spec(1, width = "3cm") |>
  # column_spec(2, width = "9cm") |>
  # column_spec(3, width = "2.5cm") |>
  row_spec(0, bold = TRUE)

```

## Statistical analysis

### Accelerated Failure Time (AFT) models

The primary outcome of interest was the age at vaccination, measured as the number of days from birth until receipt of the vaccine under consideration. For a specific antigen, let $T_i$ denote the age at vaccination for child $i$. Since vaccination status differed across children at the time of interview, three forms of observations were present. For children whose vaccination date was recorded on the vaccination card, the event time $T_i$ was observed exactly. For children whose mothers reported that the child had received the vaccine but the exact vaccination date was unavailable (typically because no vaccination card was available), the event time was interval censored such that, if $L_i$ denotes the recommended age for the vaccine (Table 1.1) and $U_i$ denotes the child's age at interview, then the true vaccination time satisfies $L_i < T_i \le U_i$. Children who had not received the vaccine by the interview date were treated as right censored, such that $T_i > U_i$. The age at vaccination was then modelled using a parametric accelerated failure time (AFT) model assuming a Weibull distribution:

$$T_i \sim \text{Weibull}(\lambda_i,\kappa),$$

where $\kappa$ denotes the Weibull shape parameter and $\lambda_i$ is the individual-specific scale parameter. The AFT model assumes a log-linear relationship between the expected survival time and the explanatory variables. For the heatwave exposure variable:

$$\log(\lambda_i)=\beta_0+\beta_1H_i+\mathbf{X}_i^\top\boldsymbol{\gamma},$$

where $\beta_0$ is the intercept, $H_i$ is the binary heatwave exposure indicator (1 = exposed, 0 = not exposed), $\mathbf{X}_i$ is the vector of adjustment covariates, and $\boldsymbol{\gamma}$ is the corresponding vector of regression coefficients.

For the Heat Index exposure variable, a second model replaced the binary heatwave indicator with the categorical heat index exposure variable:

$$\log(\lambda_i)
=
\beta_0
+
\sum_{k=2}^{K}\beta_kI(HI_i=k)
+
\mathbf{X}_i^\top\boldsymbol{\gamma},$$

where $HI_i$ denotes the heat index category for child $i$, $I(HI_i=k)$ is an indicator function for the $k^{\text{th}}$ heat index category, and the lowest heat category (Tolerable Heat Index) served as the reference level.

For an individual child, the likelihood contribution depended on the censoring mechanism. Let $f(\cdot)$, $F(\cdot)$ and $S(\cdot)=1-F(\cdot)$ denote the Weibull probability density, cumulative distribution and survival functions, respectively. The individual likelihood contribution was constructed as follows:

$$L_i(\boldsymbol{\theta})=
\begin{cases}
f(T_i), & \text{if the vaccination date was observed},\\
F(U_i)-F(L_i), & \text{if the vaccination time was interval censored},\\
S(U_i), & \text{if the child was right censored}.
\end{cases}$$

The full likelihood was then obtained by taking the product of the individual likelihood contributions,

$$L(\boldsymbol{\theta})=\prod_{i=1}^{n}L_i(\boldsymbol{\theta}),$$

where $\boldsymbol{\theta}$ denotes the vector of Weibull and regression parameters. Maximum likelihood estimates were obtained using the `svysurvreg()` function from the `survival` package in R, which fits Weibull accelerated failure time models while accommodating exact, interval-censored and right-censored observations within a unified likelihood framework that handles the multi-stage sampling design with sampling weights. Regression coefficients from the AFT model were exponentiated and interpreted as event time ratios. An event time ratio greater than one indicates a longer expected time to vaccination (greater delay), whereas an event time ratio less than one indicates a shorter expected time to vaccination (earlier vaccination), after adjusting for the selected covariates.

### Model-standardized predicted vaccination timeliness 

The model-standardized vaccination timeliness under each heat-exposure scenario was estimated for each fitted AFT model. The aim was to estimate the proportion of children who would be expected to have received the vaccine by a given number of days after the recommended vaccination date if all children in the study population were assigned the same exposure scenario, while retaining their observed covariate distributions.

For observation $i$, the fitted model gives a linear predictor $\eta_i = X_i\beta$ and its corresponding shape parameter $\lambda_i = \text{exp}(\eta_i)$; The predicted probability that child $i$ has received the vaccine by time $t$ is therefore:

$$
P(T_i \leq t | X_i) = 1-\text{exp}(-(\frac{t}{\lambda_i})^\kappa)
$$

For each exposure scenario $a$ (Heatwave/No Heatwave for the heatwave exposure, and Tolerable/Caution/Extreme Caution for the heat index exposure), the exposure variable was set to $a$ for every child, while all other observed covariates were retained. The model-standardized predicted coverage at time $t$ was then calculated as the survey-weighted average of the individual predicted probabilities:

$$
\hat{P_a}(T \leq t) = \frac{\sum_{i=1}^nw_i\cdot\hat{P}(T_i \leq t | X_i, A_i=a)}{\sum_{i=1}^nw_i}
$$

where $w_i$ denotes the survey weight for child $i$.

Thus, the resulting estimate represents the expected vaccination coverage by $t (t=7,14,21,28)$ days after the recommended vaccination date under a specified heat-exposure scenario, standardized to the covariate distribution of the study population.

### Statistical software

All analyses were conducted using R version 4.5.1[@rcoreteam2025]. Data wrangling, manipulation and visualization were performed using the `tidyverse` suite of packages[@wickham2019]. Data retrieval and management were carried out using `ecmwfr`[@hufkens2019a], `haven`[@wickham2025], and `terra`[@hijmans2026], while spatial processing and analysis were performed using the `terra` and `sf` packages[@hijmans2026; @pebesma2023]. Extreme value modelling was undertaken using `gamlss` and `gamlssx`[@northrop2025a; @r.a.rigby2005a], computation of the heat index was done using the `weathermetrics` package[@anderson2013] and model fitting was conducted using the `survey` [@lumley2004; @lumley2010] package. Reproducible reports and tables were generated using `rmarkdown` and `kableExtra`[@allaire2026; @zhu2024].

The R codes for this study are available at: https://https://github.com/stanleyrazor/Msc-VaxTimeliness-HeatWaves
